Master's Project Journal

### 2019-04-13 thesis/FPGABasedADPLLs.tex
end of part2

While it is advantageous for the purposes of analysis that this design is affected by variation in temperature, implementation and other non-idealities this brings with it challenges not present in a true mixed-signal design. In a design intended for an \ac{ASIC} the design has near total control over the placement of circuit elements, however, the \ac{FPGA} based counterpart control is available over just the area in which the oscillator may lie, and the \ac{EDA} is responsible for the exact placement. Unlike the \ac{ASIC} implementation, in which the propagation delay through each element of the ring is well bounded, the lack of control over placement means that the propagation time through each inverter is not well bounded and may vary wildly. This makes it time consuming to ensure that the frequency tuning ranges of each oscillator match that desired, and use of this oscillator less straight-forward than that of the previous two designs.

@@

There are three situations that may occur while the \ac{FSM} is in the waiting state. Firstly the case where the reference signal sees a rising edge before the generated clock, and thus the generated clock is said to be lagging the reference. As previously mentioned the convention dictates that the phase difference have a positive sign, so the counter is incremented to count in the positive direction. Similarly if the generated signal is leading the reference the sign of the error is negative and thus the counter is instructed to count downwards. The third potential scenario occurs when both clock signals have a rising edge within the same period of the \ac{FPGA} clock, which due to quantisation will be interpreted as both signals occurring at exactly the same time. In this case the counter receives no instruction and a zero value is stored until the next measurement interval. Once the \ac{FSM} is in a measurement state, and the counter either incrementing or decrementing, a rising edge occurring on the opposite signal to that which initiated the measurement will end the measurement interval.

%TODO state trans diagram

@@

To determine the sign itself, which clock edge first had a rising edge must be recorded, and this can be done by setting a \ac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle however, it is apparent that all 4 combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. Careful design of the system can avoid problems arising from this. Examining the truth table in Figure \ref{fig:sr} (a), the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Figure \ref{fig:sr} (b) illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may in other systems be feasible to dismiss this occurrence as a low probability event, the goal of a \ac{PLL} is to reduce the time delay between rising edges to zero which increases the possibility of this occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in an \ac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is however a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on an \ac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on an \ac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
%TODO sr truth table
%TODO image of simplified detector

\subsubsection{Tapped Delay Line}
To complement the asynchronous sign detection circuit, an asynchronous \ac{TDC} is also required, as otherwise there would be no benefit to using the above method over the more easily understood \ac{FSM} based design as quantisation to the clock period would happen in the \ac{TDC} rather than the sign detection circuitry. The standard method for implementing a \ac{TDC} on an \ac{ASIC} is through a tapped delay line, where a signal is propagated through a chain of fixed length delays, and at the end of the measurement interval it is recorded whether or not the signal has propagated to each point in the chain. This architecture is illustrated in Figure \ref{fig:tdl}. As with the mixed signal circuitry previously addressed, a \ac{TDL} cannot be exactly replicated on an \ac{FPGA}, but through the use of inverters it is possible to replace the unrealisable well bounded delays. As with the \ac{RO}, the delay though the inverters is not nearly as well bounded, and significant variation between delays, in both individual \ac{TDL}s and within the same \ac{TDL}, due to the implementation occurs This variation notwithstanding, the inverter pairs are a suitable replacement, and this method of time-to-digital conversion provides both resolution advantages and the presence of implementation based variation over a clocked design. These prove particularly beneficial when attempting to verify theoretical results.

Similarly to the sign at the output of the detection circuit, flip-flops store the value of each tap between phase error measurements. Prior to storage, the tap values must be converted to two's complement. As the signal propagates through the taps, magnitude of the phase error is determined by how many taps the signal has passed through. Conceptually, this thermometer coding is first converted to an unsigned integer before conversion to a signed quantity based on the sign of the error.
%TODO high level TDL image (using Z-1s)
%TODO eugene re: why no zero output

>>>>

While it is advantageous for the purposes of analysis that this design is affected by variation in temperature, implementation and other non-idealities this brings with it challenges not present in a true mixed-signal design. In a design intended for an \ac{ASIC} the design has near total control over the placement of circuit elements, however, the \ac{FPGA} based counterpart control is available over just the area in which the oscillator may lie, and the \ac{EDA} is responsible for the exact placement. Unlike the \ac{ASIC} implementation, in which the propagation delay through each element of the ring is well bounded, the lack of control over placement means that the propagation time through each inverter is not well bounded and may vary wildly. This makes it time consuming to ensure that the frequency tuning ranges of each oscillator match that desired, and use of this oscillator less straight-forward than that of the previous two designs. This is compounded by the simulation tools in \ac{EDA}s intended for \ac{FPGA}s being unable to accurately simulate the time delays, thus negating any benefit of test benches for the alignment process.

@@

There are three situations that may occur while the \ac{FSM} is in the waiting state. Firstly the case where the reference signal sees a rising edge before the generated clock, and thus the generated clock is said to be lagging the reference. As previously mentioned the convention dictates that the phase difference have a positive sign, so the counter is incremented to count in the positive direction. Similarly if the generated signal is leading the reference the sign of the error is negative and thus the counter is instructed to count downwards. The third potential scenario occurs when both clock signals have a rising edge within the same period of the \ac{FPGA} clock, which due to quantisation will be interpreted as both signals occurring at exactly the same time. In this case the counter receives no instruction and a zero value is stored until the next measurement interval. Once the \ac{FSM} is in a measurement state, and the counter either incrementing or decrementing, a rising edge occurring on the opposite signal to that which initiated the measurement will end the interval.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../state_trans_new}
	\caption[Example State Transition Diagram for a Moore Machine]{Example State Transition Diagram for a Moore Machine.}
	\label{fig:state_trans}
\end{figure}

@@

To determine the sign itself, which clock edge first had a rising edge must be recorded, and this can be done by setting a \ac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle however, it is apparent that all 4 combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. Careful design of the system can avoid problems arising from this. Examining the truth table in Table \ref{table:sr_tt}, the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Table \ref{table:sr_tt} also illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.
\begin{table}[ht]
	\begin{center}
		\begin{tabular}{cc|cc}           
			$S$&$R$&$Q$&$\overline{Q}$\\
			\hline
			0&0&$Q_{n-1}$&$\overline{Q}_{n-1}$\T\\
			0&1&0&1\T\\
			1&0&1&0\T\\
			1&1&\multicolumn{2}{c}{undefined}\T\\					
		\end{tabular}
		\caption{SR Latch Truth Table}
		\label{table:sr_tt}
	\end{center}
\end{table}

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may in other systems be feasible to dismiss this occurrence as a low probability event, the goal of a \ac{PLL} is to reduce the time delay between rising edges to zero which increases the possibility of this occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in an \ac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is however a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on an \ac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on an \ac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
\begin{table}[!ht]
	\begin{center}
		\begin{tabular}{l|cc|cc|c|cc|c}           
			When&$Gen$&$Ref$&$\overline{Gen}/S^a$&$\overline{Ref}/R^a$&$Q^a$&$S^b$&$R^b$&$Sign/Q^b$\\
			\hline
			Between Measurements&0&0&1&1&\multicolumn{1}{c|}{undefined}&\multicolumn{2}{c|}{01 or 10}&\multicolumn{1}{c}{0 or 1}\T\\
			Ref. First&0&1&1&0&1&0&1&0\T\\
			Gen. First&1&0&0&1&0&1&0&1\T\\	
			Measurement Complete&1&1&0&0&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$\T\\					
		\end{tabular}
		\caption{Sign Detector Truth Table}
		\label{table:sign_tt}
	\end{center}
\end{table}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_sign_detection}
	\caption[Example Sign Detector]{Example Sign Detector.}
	\label{fig:arbitration}
\end{figure}

\subsubsection{Tapped Delay Line}
To complement the asynchronous sign detection circuit, an asynchronous \ac{TDC} is also required, as otherwise there would be no benefit to using the above method over the more easily understood \ac{FSM} based design as quantisation to the clock period would happen in the \ac{TDC} rather than the sign detection circuitry. The standard method for implementing a \ac{TDC} on an \ac{ASIC} is through a tapped delay line, where a signal is propagated through a chain of fixed length delays, and at the end of the measurement interval it is recorded whether or not the signal has propagated to each point in the chain. This architecture is illustrated in Figure \ref{fig:simple_tdc}. As with the mixed signal circuitry previously addressed, a \ac{TDL} cannot be exactly replicated on an \ac{FPGA}, but through the use of inverters it is possible to replace the unrealisable well bounded delays. As with the \ac{RO}, the delay though the inverters is not nearly as well bounded, and significant variation between delays, in both individual \ac{TDL}s and within the same \ac{TDL}, due to the implementation occurs This variation notwithstanding, the inverter pairs are a suitable replacement, and this method of time-to-digital conversion provides both resolution advantages and the presence of implementation based variation over a clocked design. These prove particularly beneficial when attempting to verify theoretical results.

Similarly to the sign at the output of the detection circuit, flip-flops store the value of each tap between phase error measurements. Prior to storage, the tap values must be converted to two's complement. As the signal propagates through the taps, magnitude of the phase error is determined by how many taps the signal has passed through. Conceptually, this thermometer coding is first converted to an unsigned integer before conversion to a signed quantity based on the sign of the error.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_tdc}
	\caption[Example \acl{TDL} Structure]{Example \acl{TDL} Structure.}
	\label{fig:simple_tdc}
\end{figure}
%TODO eugene re: why no zero output
